Verifying Concurrent Search Structure Templates for C Programs using VST

### 2023-04-23 concurrency/CONCUR 2023/paper.tex
write more proof mechanization

\author{Duc Than Nguyen}{University of Illinois at Chicago, USA \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{William Mansky}{University of Illinois at Chicago, USA}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}

@@

@@

In the specification, the local pre-condition includes both the representation of the node $\nodeboxrep(\texttt{pn->n})$ and the resource $\texttt{R(pn->n)}$. When the \texttt{traverse} function operates on the entire search structure, it generates a subsequent node, \texttt{n'}, such that the key \texttt{k} falls within the range of node \texttt{n'}. The local post-condition is expected to encompass the representation of the new node $\nodeboxrep(\texttt{pn->n'})$, the resource $\texttt{R(pn->n)}$, and a Boolean variable, \texttt{res}, which indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not. The key to verifying the \texttt{traverse} function is identifying the loop invariant for the top-level loop. We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:
\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ (\texttt{k} \in \texttt{range})\ \land \ \nodeboxrep(\texttt{pn->n'})  \ \ast \  \texttt{R(pn->n')}   
\end{align*}
The loop invariant mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop. The \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (see line 6 in Figure \ref{traverse_lock_b}). At this point, the specification indicates that we have $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$, enabling us to prove the post-condition. On the other hand, the function terminates if \lstinline{findNext} is unable to locate a subsequent node with the key value  $\texttt{k}$ within its set of key values (refer to line 11 in Figure \ref{traverse_lock_b}). In this case, the specification clarifies that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} \neq \texttt{NULL}$ which also enables us to prove the post-condition.

The proof begins with $\nodeboxrep \texttt{(pn->n)}$, which comprises a lock invariant $\texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$, and a resource $\texttt{R(pn->n)}$ acquired before invoking the \lstinline{traverse} function. After \lstinline{findNext} successfully detects the next node (for example, \texttt{n'}) to visit, we acquire the lock of node \texttt{n'} (see line 14 in Figure \ref{traverse_lock_b}). It means we can get the resource $\texttt{R(pn->n')}$ from $\texttt{pn->n'->lock} \lockvar \texttt{R(pn->n')}$. Subsequently, we release the lock of the current node \texttt{p}. It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}). By acquiring the lock for the next node \texttt{n'} before releasing the lock for the current node \texttt{p}, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it (hand-over-hand locking). 
The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.

@@

Before proving the \texttt{traverse} function, let's define the loop invariant for the given template of the \texttt{traverse} function as follows:
\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
\end{align*}

To begin the proof, we start with the $\infp (\texttt{pn->n})$ predicate from the local pre-condition. After acquiring the lock of the node $\texttt{pn->n}$ (line 5 in Figure \ref{traverse_giveup_b}), we obtain the resource $\texttt{R(pn->n)}$ and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}. Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (see line 8 in Figure \ref{traverse_giveup_b}). At this point, we can prove the post-condition using the specification, which indicates that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$.

@@

\subsection{Lock-coupling Template}
\begin{figure}[ht]
	\begin{subfigure}[t]{0.48\textwidth}
	\lstinputlisting[language=C, style=myStyle]{lock_insert.c} 
	\caption{The \lstinline{insert} method of the lock-coupling template algorithm}
	\label{insert_lock}	
	\end{subfigure}\qquad
    \begin{subfigure}[t]{0.48\textwidth}
    \lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
    \caption{The \lstinline{insert} method of the give-up template algorithm}
    \label{insert_giveup}
    \end{subfigure} 
\caption{The \lstinline{insert} method of the lock-coupling and give-up template algorithms}
\label{insert_lock_giveup} 
\end{figure}

@@

In the \texttt{then} branch, when the \texttt{traverse} fails to find the next node, resulting in $\texttt{res} = \texttt{false}$, it implies that $\texttt{pn->n'->t} \neq \texttt{NULL}$. This indicates that the key to be inserted already exists in the tree. In this scenario, we update the node's value while maintaining the rest of the tree structure (line 6 in Figure \ref{insert_lock}); this acts as the linearization point for the insertion. Conversely, when the \texttt{traverse} reaches a leaf and produces $\texttt{res} = \texttt{true}$, it implies $\texttt{pn->n'->t} = \texttt{NULL}$ (line 9 in Figure \ref{insert_lock}). We then utilize the \texttt{insertOp} specification, and with enough information, we can conclude that inserting the key at this location will alter the abstract state from an unknown $\texttt{m}$ to $\texttt{m[k} \mapsto \texttt{v]}$. In both instances, we must prove that we can atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ (\texttt{m[k} \mapsto \texttt{v])}$, by demonstrating that the in-memory tree update corresponds to a local modification in the abstract tree, which inserts the new key-value pair. 
Next, we release (line 11) the node \texttt{n} and return it to the shared state. Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.

@@

The proof for the \texttt{insert} operation begins with the preconditions $\nodeboxrep \ \texttt{p}$ and $\treerep \ \texttt{m}$. We assign \texttt{pn->n} to point to the same memory address as \texttt{*p} (refer to line 3 in Figure \ref{insert_giveup}). In Section \ref{traverse_proof_giveup}, we discuss providing the identical preconditions for the \texttt{traverse} function, which encompasses the local precondition $\infp(\texttt{pn->n})$ and the public precondition $\treerep\ \texttt{m}$. Subsequently, we apply the \texttt{traverse} specification from Section \ref{traverse_proof_giveup}.

@@

\section{Proof Mechanization}
We present the mechanized formal verification of the correctness of these templates, which are implemented in C and verified with the Verified Software Toolchain (VST). We employ concurrent separation logic (CSL) to analyze the behavior of the C code and incorporate Iris-style logical atomicity within VST.

By utilizing VST, we can create programs in standard C, automatically generate abstract syntax trees (ASTs) for them in Coq, and subsequently state and interactively prove specifications for these programs with the support of automated tactics. All proofs are formalized in Coq using VST and can be found online at

@@

\section{Related Work}

@@

%% Please use bibtex,

\bibliography{../sources}

>>>>

\author{Duc Than Nguyen}{University of Illinois at Chicago, USA \and \url{http://dnguye96.people.uic.edu} }{dnguye96@uic.edu}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Lennart Beringer}{Princeton University, USA \and \url{https://www.cs.princeton.edu/~eberinge} }{eberinge@cs.princeton.edu}{https://orcid.org/0000-0002-1570-3492}{(Optional) author-specific funding acknowledgements}

\author{William Mansky}{University of Illinois at Chicago, USA\and \url{https://mansky.lab.uic.edu/} }{mansky1@uic.edu}{https://orcid.org/0000-0002-5351-895X}{(Optional) author-specific funding acknowledgements}

\author{Shengyi Wang}{Princeton University, USA \and \url{https://www.cs.princeton.edu/~shengyiw} }{shengyiw@cs.princeton.edu}{https://orcid.org/[0000-0002-2286-8703}{(Optional) author-specific funding acknowledgements}

@@

\paragraph*{Related Work}
\than{I think we will write related work inside introduction part}

@@

In the specification, the local pre-condition includes both the representation of the node $\nodeboxrep(\texttt{pn->n})$ and the resource $\texttt{R(pn->n)}$. When the \texttt{traverse} function operates on the entire search structure, it generates a subsequent node, \texttt{n'}, such that the key \texttt{k} falls within the range of node \texttt{n'}. The local post-condition is expected to encompass the representation of the new node $\nodeboxrep(\texttt{pn->n'})$, the resource $\texttt{R(pn->n)}$, and a Boolean variable, \texttt{res}, which indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not. The key to verifying the \texttt{traverse} function is identifying the loop invariant for the top-level loop. We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:
\begin{mathpar}\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ (\texttt{k} \in \texttt{range})\ \land \ \nodeboxrep(\texttt{pn->n'})  \ \ast \  \texttt{R(pn->n')}   
\end{mathpar}
The loop invariant mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop. The \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (line 6 in Figure \ref{traverse_lock_b}). At this point, the specification indicates that we have $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$, enabling us to prove the post-condition. On the other hand, the function terminates if \lstinline{findNext} is unable to locate a subsequent node with the key value  $\texttt{k}$ within its set of key values (line 11 in Figure \ref{traverse_lock_b}). In this case, the specification clarifies that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} \neq \texttt{NULL}$ which also enables us to prove the post-condition.

The proof begins with $\nodeboxrep \texttt{(pn->n)}$, which comprises a lock invariant $\texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$, and a resource $\texttt{R(pn->n)}$ acquired before invoking the \lstinline{traverse} function. After \lstinline{findNext} successfully detects the next node (for example, \texttt{n'}) to visit, we acquire the lock of node \texttt{n'} (line 14 in Figure \ref{traverse_lock_b}). It means we can get the resource $\texttt{R(pn->n')}$ from $\texttt{pn->n'->lock} \lockvar \texttt{R(pn->n')}$. Subsequently, we release the lock of the current node \texttt{p}. It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}). By acquiring the lock for the next node \texttt{n'} before releasing the lock for the current node \texttt{p}, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it (hand-over-hand locking). 
The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.

@@

Before proving the \texttt{traverse} function, let's define the loop invariant for the given template of the \texttt{traverse} function as follows:
\begin{mathpar} \texttt{traverse\_inv(pn)} \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
\end{mathpar}
To begin the proof, we start with the $\infp (\texttt{pn->n})$ predicate from the local pre-condition. After acquiring the lock of the node $\texttt{pn->n}$ (line 5 in Figure \ref{traverse_giveup_b}), we obtain the resource $\texttt{R(pn->n)}$ and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}. Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (line 8 in Figure \ref{traverse_giveup_b}). At this point, we can prove the post-condition using the specification, which indicates that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$.

@@

\subsection{Lock-coupling Template}
\begin{figure}[ht]
	\begin{subfigure}[t]{0.48\textwidth}
	\lstinputlisting[language=C, style=myStyle, escapechar=|]{lock_insert.c} 
	\caption{The \lstinline{insert} method of the lock-coupling template algorithm}
	\label{insert_lock}	
	\end{subfigure}\qquad
    \begin{subfigure}[t]{0.48\textwidth}
    \lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
    \caption{The \lstinline{insert} method of the give-up template algorithm}
    \label{insert_giveup}
    \end{subfigure} 
\caption{The \lstinline{insert} method of the lock-coupling and give-up template algorithms}
\label{insert_lock_giveup} 
\end{figure}

@@

In the \texttt{then} branch, when the \texttt{traverse} fails to find the next node, resulting in $\texttt{res} = \texttt{false}$, it implies that $\texttt{pn->n'->t} \neq \texttt{NULL}$. This indicates that the key to be inserted already exists in the tree. In this scenario, we update the node's value while maintaining the rest of the tree structure (line 6 in Figure \ref{insert_lock}); this acts as the linearization point for the insertion. Conversely, when the \texttt{traverse} reaches a leaf and produces $\texttt{res} = \texttt{true}$, it implies $\texttt{pn->n'->t} = \texttt{NULL}$ (line 9 in Figure \ref{insert_lock}). We then utilize the \texttt{insertOp} specification, and with enough information, we can conclude that inserting the key at this location will alter the abstract state from an unknown $\texttt{m}$ to $\texttt{m[k} \mapsto \texttt{v]}$. In both instances, we must prove that we can atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ (\texttt{m[k} \mapsto \texttt{v])}$, by demonstrating that the in-memory tree update corresponds to a local modification in the abstract tree, which inserts the new key-value pair. 
Next, we release (line 11 in Figure \ref{insert_lock}) the node \texttt{n} and return it to the shared state. Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.

@@

The proof for the \texttt{insert} operation begins with the preconditions $\nodeboxrep \ \texttt{p}$ and $\treerep \ \texttt{m}$. We assign \texttt{pn->n} to point to the same memory address as \texttt{*p} (line 3 in Figure \ref{insert_giveup}). In Section \ref{traverse_proof_giveup}, we discuss providing the identical preconditions for the \texttt{traverse} function, which encompasses the local precondition $\infp(\texttt{pn->n})$ and the public precondition $\treerep\ \texttt{m}$. Subsequently, we apply the \texttt{traverse} specification from Section \ref{traverse_proof_giveup}.

@@

\section{Proof Mechanization}
We present the mechanized formal verification of the correctness of these templates, which are implemented in C and verified with the Verified Software Toolchain. We employ concurrent separation logic to analyze the behavior of the C code and incorporate Iris-style logical atomicity within VST.

\begin{table}[h]
	\begin{center}
		\begin{tabular}{l | c || l | c } 
			\hline
			\textbf{Lock-coupling template} & \textbf{Line of proofs} & \textbf{Give-up  template} & \textbf{Line of proofs} \\
			\hline
			Template specfications & 190 & Template specfications & 138 \\
			\texttt{findNext} proof 	& 27  & \texttt{findNext} proof & 27 \\
			\texttt{insertOp} proof 	& 32  & \texttt{insertOp} proof & 31 \\
			& & \texttt{inRange} proof & 30 \\
			\texttt{traverse} proof	& 360  & \texttt{traverse} proof & 336 \\ 
			\texttt{insert} proof 	   		& 450  & \texttt{insert} proof & 284 \\
			\texttt{lookup} proof 	   	& 235  &  \texttt{lookup} proof & 304 \\ 
			\texttt{pushdown\_left} proof 	   	& 282 & & \\	
			\texttt{delete} proof 		& 182	 &   & 	 \\
			supporting proofs    & 	1909	 & supporting proofs 	& 1876 \\
			\hline
			\textbf{Total} & \textbf{3667} & & \textbf{3026}  
		\end{tabular}
		\caption{Size of Coq definitions and proofs, by topic.}
		\label{table_coq}
	\end{center}
\end{table}

By utilizing VST, we can create programs in standard C, automatically generate abstract syntax trees (ASTs) for them in Coq, and subsequently state and interactively prove specifications for these programs with the support of automated tactics. Table \ref{table_coq} presents a summary of our Coq development. We formalized all previously described proofs in Coq, taking advantage of VST's features for verifying separation logic specifications of C functions. This encompasses supplementary proofs, template specifications (encompassing all helper functions such as \texttt{findNext}, \texttt{insertOp}, and others, in addition to \texttt{traverse}, \texttt{insert}, \texttt{lookup}, and \texttt{delete}), as well as proofs against with lock-coupling and give-up template specifications. All proofs are formalized in Coq using VST and can be found online at

@@

@@

%% Please use bibtex, 
\newpage
\bibliography{../sources}
